### 2022-11-13 chapter1.tex
Added TODO notes and new sections

In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}). Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches. However, we modified many details to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, since modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the overall locus of genes, the accurate identification of exon boundaries and start codons when transcript evidence is limited remains an ongoing challenge~\cite{Frankish2015, Dunne2018}. Consequently, protein sequences derived from annotation pipelines can include non-homologous segments of significant length or exclude highly conserved segments. Such heterogeneity in the structure and length of the sequences in an orthologous group poses many challenges for their alignment and subsequent analysis. Thus, we implemented several novel quality control and data cleaning steps to correct mis-alignments and identify likely sequencing, assembly, or annotation errors.

@@

\section*{Results}
\subsection*{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. The complete list of genomes is given in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

@@

\subsection*{Clustering in orthologous groups}
The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}). Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated. Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes. Others are possibly novel genes. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data yields many nodes which form triangles but lack dense networks of edges with other nodes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\section*{Discussion}
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteome and specific gene families in the \textit{Drosophila} genus. To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}). Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.

@@

>>>>

In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant. Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches. However, we modified many details to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, since modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the overall locus of genes, the accurate identification of exon boundaries and start codons when transcript evidence is limited remains an ongoing challenge~\cite{Frankish2015, Dunne2018}. Consequently, protein sequences derived from annotation pipelines can include non-homologous segments of significant length or exclude highly conserved segments. Such heterogeneity in the structure and length of the sequences in an orthologous group poses many challenges for their alignment and subsequent analysis. Thus, we implemented several novel quality control and data cleaning steps to correct mis-alignments and identify likely sequencing, assembly, or annotation errors.

@@

\section*{Results}
\subsection*{Pipeline overview}
% TODO: Add description of pipeline, adapting text from Figure 1 caption

\subsection*{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. The complete list of genomes is given in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

@@

\subsection*{Clustering in orthologous groups}
% TODO: Add description of structure of connected components to motivate clustering algorithm
The best hits between sequences are naturally visualized as a graph where the sequences are nodes and best hits are edges between nodes (Fig.~\ref{fig:components}). Some sets of nodes form tight clusters whereas others are more loosely associated or even isolated. Automated genome annotation is not completely accurate, so some poorly supported hits are likely spurious, arising from homologous regions that were falsely annotated as genic in two closely related genomes. Others are possibly novel genes. Regardless of their origin, these hits are not strong candidates for comparative analyses since an orthology relationship is supported by relatively few genome pairs. In contrast, likely orthologs should consistently identify each other as best reciprocal hits across many genome pairs. Thus, orthologous groups can be operationally defined as self-consistent clusters in the hit graph. Since sequence divergence or assembly and annotation errors may prevent a best reciprocal hit between orthologous sequences across all genome pairs, the clustering method should require a high degree of self-consistency without demanding complete consensus.

@@

% TODO: Add more discussion of different clustering methods and their weaknesses (include citations)
The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data yields many nodes which form triangles but lack dense networks of edges with other nodes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Inference of species trees}
% TODO: Add discussion of creating species tree

@@

\section*{Discussion}
% TODO: Re-write to remove species tree and combine with following paragraph; also highlight specific contributions
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, consensus phylogenetic trees were inferred from meta-alignments sampled from alignments of single copy orthologous groups (Fig.~\ref{fig:trees}). Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Thus, the orthologous groups and the initial alignments with and without non-representative sequences are included as supplemental data.
